%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% I, the copyright holder of this work, release this work into the
%% public domain. This applies worldwide. In some countries this may
%% not be legally possible; if so: I grant anyone the right to use
%% this work for any purpose, without any conditions, unless such
%% conditions are required by law.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[
  digital,     %% The `digital` option enables the default options for the
               %% digital version of a document. Replace with `printed`
               %% to enable the default options for the printed version
               %% of a document.
%%  color,       %% Uncomment these lines (by removing the %% at the
%%               %% beginning) to use color in the printed version of your
%%               %% document
  oneside,     %% The `oneside` option enables one-sided typesetting,
               %% which is preferred if you are only going to submit a
               %% digital version of your thesis. Replace with `twoside`
               %% for double-sided typesetting if you are planning to
               %% also print your thesis. For double-sided typesetting,
               %% use at least 120 g/m² paper to prevent show-through.
  nosansbold,  %% The `nosansbold` option prevents the use of the
               %% sans-serif type face for bold text. Replace with
               %% `sansbold` to use sans-serif type face for bold text.
  nocolorbold, %% The `nocolorbold` option disables the usage of the
               %% blue color for bold text, instead using black. Replace
               %% with `colorbold` to use blue for bold text.
  lof,         %% The `lof` option prints the List of Figures. Replace
               %% with `nolof` to hide the List of Figures.
  lot,         %% The `lot` option prints the List of Tables. Replace
               %% with `nolot` to hide the List of Tables.
]{fithesis4}
%% The following section sets up the locales used in the thesis.
\usepackage[resetfonts]{cmap} %% We need to load the T2A font encoding
\usepackage[T1,T2A]{fontenc}  %% to use the Cyrillic fonts with Russian texts.
\usepackage[
  main=english, %% By using `czech` or `slovak` as the main locale
                %% instead of `english`, you can typeset the thesis
                %% in either Czech or Slovak, respectively.
  english, german, russian, czech, slovak %% The additional keys allow
]{babel}        %% foreign texts to be typeset as follows:
%%
%%   \begin{otherlanguage}{german}  ... \end{otherlanguage}
%%   \begin{otherlanguage}{russian} ... \end{otherlanguage}
%%   \begin{otherlanguage}{czech}   ... \end{otherlanguage}
%%   \begin{otherlanguage}{slovak}  ... \end{otherlanguage}
%%
%% For non-Latin scripts, it may be necessary to load additional
%% fonts:
\usepackage{paratype}
\def\textrussian#1{{\usefont{T2A}{PTSerif-TLF}{m}{rm}#1}}
%%
%% The following section sets up the metadata of the thesis.
\thesissetup{
    date        = 2024/04/02, % TODO date
    university  = mu,
    faculty     = fi,
    type        = mgr,
    department  = Department of Computer Systems and Communications,
    author      = Bc. Martin Bartoš,
    gender      = m,
    advisor     = {prof. RNDr. Václav Matyáš, M.Sc., Ph.D.},
    title       = {Adaptive Authentication in Keycloak},
    TeXtitle    = {Adaptive Authentication in Keycloak},
    keywords    = {adaptive, authentication, Keycloak, open-source, Red Hat, risk-based, security},
    TeXkeywords = {adaptive, authentication, Keycloak, open-source, Red Hat, risk-based, security},
    abstract    = {%
      Keycloak is an open-source identity and access management system that provides robust security solutions for applications and services. It acts as a centralized authentication and authorization platform, enabling organizations to secure their resources while offering Single Sign-On capabilities. Keycloak is widely used in various industries, including web applications, microservices, and APIs, to protect sensitive data and streamline user access control. 
      
      The thesis aims to enhance the security of authentication processes in Keycloak. Traditional static authentication mechanisms, such as username and password, are vulnerable to various security threats, including password breaches, phishing attacks, and unauthorized access. Adaptive authentication might help mitigate these issues. 
      
      Adaptive authentication aims to improve the security posture by dynamically adjusting the authentication requirements based on contextual factors. That means the system can adapt its authentication methods and policies in response to changing circumstances.
    },
    thanks      = {%
      I am deeply grateful to my supervisor, Václav Matyáš, for his guidance and support throughout this thesis. I would also like to thank Marek Posolda for his valuable insights, which greatly enriched this work. Their contributions have been instrumental in its completion.
    },
    bib         = adaptive_authn.bib,
    %% Remove the following line to use the JVS 2018 faculty logo.
    facultyLogo = fithesis-fi,
}
\usepackage{makeidx}      %% The `makeidx` package contains
\makeindex                %% helper commands for index typesetting.
%% These additional packages are used within the document:
\usepackage{paralist} %% Compact list environments
\usepackage{amsmath}  %% Mathematics
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{url}      %% Hyperlinks
\usepackage{markdown} %% Lightweight markup
\usepackage{listings} %% Source code highlighting
\usepackage{graphicx}
\usepackage{rotating}
\lstset{
  basicstyle      = \ttfamily,
  identifierstyle = \color{black},
  keywordstyle    = \color{blue},
  keywordstyle    = {[2]\color{cyan}},
  keywordstyle    = {[3]\color{olive}},
  stringstyle     = \color{teal},
  commentstyle    = \itshape\color{magenta},
  breaklines      = true,
}
\usepackage{floatrow} %% Putting captions above tables
\floatsetup[table]{capposition=top}
\usepackage[babel]{csquotes} %% Context-sensitive quotation marks
\usepackage{import}

\begin{document}
%% The \chapter* command can be used to produce unnumbered chapters:
\chapter*{Introduction}
%% Unlike \chapter, \chapter* does not update the headings and does not
%% enter the chapter to the table of contents. I we want correct
%% headings and a table of contents entry, we must add them manually:
\markright{\textsc{Introduction}}
\addcontentsline{toc}{chapter}{Introduction}

<TODO>
In today's digital landscape, where cybersecurity threats continue to evolve in sophistication and frequency, the need for robust authentication mechanisms to safeguard sensitive data and systems has become paramount.
Traditional authentication methods, such as static passwords or even static multi-factor authentication (MFA), often fall short of adequately addressing the dynamic nature of security risks and user behavior.
As a response to these challenges, adaptive authentication has emerged as a promising approach to authentication, offering a dynamic and context-aware method for verifying user identities.

This thesis aims to explore the principles, mechanisms, benefits, and challenges of adaptive authentication in-depth.
By examining the underlying concepts and technologies, as well as conducting case studies and analyses, this research seeks to provide insights into the effectiveness and applicability of adaptive authentication solutions in diverse organizational contexts.
Through a comprehensive examination of adaptive authentication, this thesis aims to contribute to the understanding of modern authentication practices and their role in mitigating cybersecurity risks.

\shorthandoff{-}
\begin{markdown*}{%
  hybrid,
  definitionLists,
  footnotes,
  inlineFootnotes,
  hashEnumerators,
  fencedCode,
  citations,
  citationNbsps,
  pipeTables,
  tableCaptions,
}

\import{sections/}{1-adaptive-authentication.tex}
\import{sections/}{2-existing-solutions.tex}
\import{sections/}{3-keycloak.tex}
\import{sections/}{4-specification.tex}
\import{sections/}{5-design.tex}
\import{sections/}{6-implementation.tex}

\end{markdown*}
\shorthandon{-}

\printbibliography[heading=bibintoc]

\appendix %% Start the appendices.
\chapter{An appendix}

\end{document}
